The role of aneuploidy in the evolution of cancer drug resistance

### 2024-04-03 EvolutionaryRescue1.tex
Add files via upload

Here we build on previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We wish to understand what effect does aneuploidy have on the probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells. We also calculate the mean time that an initially declining tumor cell population reaches its pre-treatment size. Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.

@@

We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could be evolutionarily rescued even without aneuploidy.
For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 89\%.
Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

@@

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploidy is likely produced even before onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}. Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but no the division rate and therefre use $\lambda_w$ for the wildtype division rate in the drug-free enviornment).

Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost $s$, such that
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a strong effect of the drug on sensitive cells, or if the cost aneuploidy cost in the drug-free environment, $s$, is small.
In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=2\times10^{-3}, s=0.07$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 8.75$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for direct mutation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.  In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large. The ratio does not depend on the rate of chromosome missegregation induced by the drug $u$. However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.9625$, which means that standing genetic variation reduces the threshold tumor size by approximately $4\%$. Therefore, standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, but it does offer a slight advantage when compared with direct mutation.

@@

\paragraph{Evolutionary rescue time.}
We have derived approximations for $\tau_m$, the mean evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the mean rescue time with aneuploidy ($u>0$), both conditioned on evolutionary rescue occurring.
These approximations are in good agreement with simulation results for small, intermediate, and large tumor sizes (\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot}).
The mean rescue time with aneuploidy for small and large tumors follows these expressions (Appendix \ref{sec:appendix_rescue_time}),
\begin{equation}  \label{eq:AsymptoticTimeRules}
\tau_a \approx \begin{cases}
    \frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{v\lambda_w N}   \frac{\lambda_m}{\Delta_m} ,& % I replaced pm with 𝜆𝑚 − 𝜇𝑚 / 𝜆𝑚
  N \gg N_m^* .
  \end{cases}
\end{equation}

@@

When a fraction $f\approx 0.284\%$ of the initial cancer cell population is aneuploid we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation. For this purpose we calculate  $\tau_a$ (see \cref{meantimet2SGV}) and compare our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}. We note that standing genetic variation does not have a significant effect on the mean evolutionary rescue time. % WHAT is the purpose of this sentence? ok

In Appendix E we derive the probability that by time $t$ a successful mutant has been generated. \Cref{cdffig}A show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent. We observe that aneuploidy starts to have an effect on adaptation for timescales greater then $1/\Delta_a\approx100$ days after which no more rescue mutations are generated through direct mutation. This shows that rescue mutations are generated through aneuploidy at latter timescales then direct mutation and, as a consequence, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.   % WHAT is the purpose of this sentence? ok

@@

We next approximate the mean time for the population of mutant cancer cells to reach the initial, pre-treatment population size $N$, which we denote the recurrence time $\tau_a^r$ (Appendix \ref{sec:appendix_recurrence_time}),
\begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
\tau_a^r \approx \begin{cases}
   \frac{1}{ \abs{\Delta_w}}+\frac{1}{ \abs{\Delta_a}}+\frac{\log p_mN}{\Delta_m} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
\Cref{proliferationFigure,ProliferationTimeLarge} shows the agreement between our approximations and simulations.
For small tumors ($N \ll N_a^*$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$. The mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time. Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates, $\Delta_w$ and $\Delta_a$ respectively, increases the recurrence time.
For large tumors ($N \gg N_m^*$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes constant (and independent of the initial tumor size $N$).
Increasing either the mutant growth rate $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time for the tumor to rebound to its initial size.
In addition, drugs that significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delay cancer relapse.
Consequently, patients treated with such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.

We notice that, for small and large tumors, the asymptotic expressions for the mean recurrence time are independent of the chromosome missagregation rate $u$ and, as a result, the rate at which the chemotherapeutic drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$ when $N \ll N_a^*$ or $N \gg N_m^*$.
% Remus the next paragraphs look like you wanted to find approximations so you did, you reference them here, but the results are buried in the technical text: WHAT IS IT THAT YOU WANT US TO LEARN from this? ok

In Appendix F we derive the probability that mutant cancer cell population has not reached size $N$ at time $t$. \Cref{cdffig}B shows agreement between our approximations and stochastic simulations for various values of $N$. Additionally, we derived the distribution of the recurrence time and plotted our result in \Cref{KMdistribution} alongside simulations for the case $N=10^6$ (small tumor) and we note that the distribution is wide and rights-skewed. It is highly unlikely to observe the recurrence of tumors at times smaller then $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}\approx1542$ days for the parameter values $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ and independent of initial tumor size $N$ (\Cref{cdffig}B).

The recurrence can also be defined as the time necessary for the tumor size to reach a detection threshold $M$. We derive the mean recurrence time to detection size $M=10^7$ in appendix \ref{sec:appendix_recurrence_time} and plot in \Cref{RecurrencePlot} with simulations. We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{r,M}$ is negligible (i.e. $\tau_a^r\approx\tau_a^{r,M}$ for $N<N_m^*$). % WHAT DOES THAT MEAN? ok
However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$.

Most clinical trials report data on the distribution of recurrence time measured from the time of surgery~\citep{avanzini2019cancer} with chemotherapy usually following after. % WHAT DOES "varying" mean here? ok
Therefore, since only undetected secondary tumors are present at the time of the administration of the anti-cancer drug we lack knowledge of the size of the tumors and we cannot compare the empirical distributions to our predictions. However, we expect that the variability of secondary tumors to average out across large cohorts of patients and, as a result, we can use the mean recurrence time. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes the population to decline in size towards potential extinction.
In this scenario, the tumor can be "evolutionary rescued", or escape extinction, via two paths. In the direct path, a sensitive cell acquires a mutation that confers resistance that allows it to rapidly grow. In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).

Using multitype branching processes, we derived the probability of evolutionary rescue of the population of cancer cells under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.

\paragraph{Evolutionary rescue}
When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step process in which aneuploidy itself rescues the population (\Cref{rescue_threshold}A). 
When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population declines slower then the sensitive cell population (\Cref{sampleTrajectories}). In this case, aneuploidy provides two benefits. First, it delays the extinction of the population--providing more time for appearance of the resistance mutation. Second, it increases the population size relative to a sensitive population--providing more cells in which mutations can occur, i.e., it increases the mutation supply, $Nuv\lambda_w\lambda_a/\abs{\Delta_w\Delta_a}$.

We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least one order of magnitude even when aneuploidy is tolerant (\Cref{rescue_prob}). Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}. This can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}. This is reinforced by the fact that metastases have been shown to have higher chromosome missagregation rate two to three orders of magnitude higher then primary tumors~\citep{kimmel2023intra}.

Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 1000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A),
this can explain the reappearance of cancer even after initial remission. The theoretical predictions for the mean rescue time for tumors smaller then  $10^8$ is greater then 4 years which is consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.

We hypothesized that presence of \emph{standing variation} 
--the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. However, we observe that for reasonable parameter values evolutionary rescue is more likely to occur through de-novo aneuploidy (\Cref{rescue_denovo}). If the the fraction of the tumor cells which have the ``beneficial'' aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$ then evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells. In that case, evolutionary rescue occurs through a single step by the aneuploid cells acquiring a beneficial mutation and the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$  (\Cref{rescue_prob_sgv}).

\paragraph{Experimental future direction}
Our model's predictions may be tested by experiments \citep{martin2013probability}. For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes. Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy or to saline solution for control~\citep{ippolito2021gene}. 
Cell density can then be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case with no cells present. We compare to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go extinct.

Additionally, we can test our model's predictions with data from patients. Given the type of cancer they have and the degree of ``beneficial'' aneuploidy,  we can obtain the parameters for our model and predict with what probability will the cancer relapse after treatment.

>>>>

Here we build on previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We wish to understand what effect does aneuploidy have on the probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells. We also calculate the mean time that an initially declining tumor cell population reaches its pre-treatment size. Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics. Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.

@@

We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment even without aneuploidy.
For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 89\%.
Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

@@

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploidy is likely produced even before onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}. Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but no the division rate and therefore use $\lambda_w$ for the wildtype division rate in the drug-free environment).

Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost $s$, such that
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
Standing genetic variation will drive evolutionary rescue if wildtype growth rate $\Delta_w$ is very negative due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $s$, is small.  
In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=2\times10^{-3}, s=0.07$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 8.75$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via direct mutation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.  In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large. The ratio does not depend on the rate of chromosome missegregation induced by the drug $u$. However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.9625$, which means that standing genetic variation reduces the threshold tumor size by approximately $4\%$. Therefore, standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, but it does offer a slight advantage when compared with direct mutation.

@@

\paragraph{Evolutionary rescue time.}
In Appendix \ref{sec:appendix_rescue_time} we have derived approximations for $\tau_m$, the mean evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the mean rescue time with aneuploidy ($u>0$), both conditioned on evolutionary rescue occurring.
These approximations are in good agreement with simulation results for small, intermediate, and large tumor sizes (\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot}).
The mean rescue time with aneuploidy for small and large tumors follows these expressions (Appendix \ref{sec:appendix_rescue_time}),
\begin{equation}  \label{eq:AsymptoticTimeRules}
\tau_a \approx \begin{cases}
    \frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{v\lambda_w N}   \frac{\lambda_m}{\Delta_m} ,& % I replaced pm with 𝜆𝑚 − 𝜇𝑚 / 𝜆𝑚
  N \gg N_m^* .
  \end{cases}
\end{equation}

@@

Given that a fraction $f\approx 0.284\%$ of the initial cancer cell population is expected to be aneuploid even before the drug administration we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation. For this purpose we calculate the mean evolutionary rescue time with standing genetic variation $\tau_a^f$ (see \cref{meantimet2SGV}) and compare our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}. We note that standing genetic variation does not have a significant effect on the mean evolutionary rescue time. % WHAT is the purpose of this sentence? ok

In Appendix \ref{sec:appendix_distribution_time} we calculate the probability that a successful mutation, which will rescue the population, has been generated by time $t$. This allows us to observe whether aneuploidy accelerates or delays adaptation. We plot our results in  \Cref{cdffig}A alongside simulations for different aneuploid growth rates ($u>0$) and the case when aneuploidy is absent ($u=0$). We observe that aneuploidy starts to have an effect on adaptation for timescales greater then $1/\Delta_a\approx100$ days after which no more rescue mutations are generated through direct mutation. This shows that rescue mutations are generated through aneuploidy at latter timescales then direct mutation and, as a consequence, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.   % WHAT is the purpose of this sentence? ok

@@

We next approximate the mean time for the population of mutant cancer cells to reach the initial, pre-treatment population size $N$, which we denote the recurrence time $\tau_a^r$ (Appendix \ref{sec:appendix_recurrence_time}),
\begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
\tau_a^r \approx \begin{cases}
   \frac{1}{ \abs{\Delta_w}}+\frac{1}{ \abs{\Delta_a}}+\frac{\log p_mN}{\Delta_m} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
\Cref{proliferationFigure,ProliferationTimeLarge} show the agreement between our approximations and simulations.
For small tumors ($N \ll N_a^*$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for its lineage to reach size $N$. The mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time. Increasing the mutant growth rate $\Delta_m$ decreases recurrence times while increasing the wildtype and aneuploid growth rates, $\Delta_w$ and $\Delta_a$ respectively, increases the recurrence time.
For large tumors ($N \gg N_m^*$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes constant (and independent of the initial tumor size $N$).
Increasing either the mutant growth rate $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time for the tumor to rebound to its initial size.
In addition, drugs that significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delay cancer recurrence.
Consequently, patients treated with such drugs may require a longer period of monitoring to guarantee the effectiveness of the treatment.

We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missagregation rate $u$, therefore the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.
% Remus the next paragraphs look like you wanted to find approximations so you did, you reference them here, but the results are buried in the technical text: WHAT IS IT THAT YOU WANT US TO LEARN from this? ok

In Appendix F we derive the probability that a mutant cancer cell population has not reached size $N$ by time $t$. \Cref{cdffig}B shows agreement between our approximations and stochastic simulations for various values of $N$. Additionally, we derive the distribution of the recurrence time for the case $N=10^6$ (small tumor), noting that the distribution is wide and right-skewed (\Cref{KMdistribution}). It is highly unlikely to observe the recurrence of tumors at times smaller then $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}\approx1542$ days for the parameter values $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ and independent of initial tumor size $N$~(\Cref{cdffig}B).

The detection time $\tau_a^{r,M}$ can be defined as the time necessary for the tumor size to reach detection threshold $M$. We derive the mean recurrence time to detection size $M=10^7$ in appendix \ref{sec:appendix_recurrence_time}. We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{r,M}$ is negligible when compared to the case where the detection size is $N$ (i.e. $\tau_a^r\approx\tau_a^{r,M}$ for $N<N_m^*$). % WHAT DOES THAT MEAN? ok
However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$ while $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}). Additionally, for large tumors we have $M<N_m^*<N$ which tells us that the mean detection time is smaller for such tumors when compared to the mean recurrence time.

Most clinical trials report data on the distribution of recurrence time measured from the time of surgery~\citep{avanzini2019cancer} with drug therapy usually following after. % WHAT DOES "varying" mean here? ok
Therefore, since only undetected secondary tumors are present at the time of the administration of the anti-cancer drug we lack knowledge of the size of the tumors and we cannot compare the empirical distributions to our predictions. However, we expect that the variability of secondary tumors to average out across large cohorts of patients, thus we can use the mean recurrence time to compare with clinical data. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes it to decline in size towards potential extinction.
In this scenario, the tumor can be "evolutionary rescued", or escape extinction, via two paths. In the direct path, a sensitive cell acquires a mutation that confers resistance that allows it to rapidly grow. In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).

\paragraph{Evolutionary rescue}
Using multitype branching processes, we derived the probability of evolutionary rescue of the tumor under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.

When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), aneuploidy itself rescues the population (\Cref{rescue_threshold}A). 
When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population size declines slower then that of the sensitive cell population (\Cref{sampleTrajectories}). In this case, aneuploidy provides two benefits. First, it delays the extinction of the population, providing more time for appearance of the resistance mutation. Second, it increases the population size relative to a sensitive population, providing more cells in which mutations can occur, i.e., it increases the mutation supply (i.e. $Nuv\lambda_w\lambda_a/\abs{\Delta_w\Delta_a}$).

We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least one order of magnitude even when aneuploidy is tolerant (\Cref{rescue_prob}). Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough (i.e. $N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}, and this can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}. The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missagregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.

Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 1000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A),
aneuploidy can explain the reappearance of cancer even after initial remission. The theoretical prediction for the mean rescue time for tumors smaller then  $10^8$ cells is greater then 4 years, consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}. We observe from \Cref{cdffig}A that aneuploidy complements evolutionary rescue through direct mutation given that aneuploidy is most likely to generate rescue mutations at time scales larger then the timescales at which rescue mutations are generated through direct mutation.

We hypothesized that \emph{standing variation} (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. From \Cref{eq:Nsgv} we observe that a drug or combination of drugs that reduces the wildtype growth rate and does not significantly increase the chromosome missegregation rate is more likely to cause evolutionary rescue to occur through direct mutation. Furthermore, we find that for reasonable parameter values evolutionary rescue is more likely to occur through \emph{de-novo} aneuploidy (\Cref{rescue_denovo}). If the fraction of tumor cells that have the ``advantageous'' aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation, rather then through \emph{de novo} aneuploid cells. In this case, evolutionary rescue likely occurs via aneuploid cells that acquire a resistance mutation and the probability of evolutionary rescue declines exponentially with initial tumor size  (the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$, see \Cref{rescue_prob_sgv}).

\paragraph{Experimental future direction}
Our model prediction could be tested by experiments \citep{martin2013probability}. For example, to assess the effect of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes. Then, these tumors may be exposed to anti-cancer drugs that induce aneuploidy or to saline solution for control~\citep{ippolito2021gene}. 
Cell density can be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case with no cells present. We can then compare the results of the experiments to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go killed by the drug.

Additionally, our model predictions can be tested with data from patients. Given the type of cancer and how ``advantageous'' the aneuploidy is,  our model can predict the probability for cancer relapse after treatment.
